Osa

### 2021-09-02 article/article.Rmd
fixed bugs, revisiting framework

Forested areas in the tropics offer many [economic and regulatory] services like timber, fuel, biomass/carbon storage, and biodiversity maintenance. Today, most forested areas in the tropics are secondary (FAO 2020) and show legacy effects of human activity like logging and intensive agriculture.

Previous studies show some consistent patterns of secondary forest succession, while others show inconsistent trajectories. For example, syntheses (Finegan 1996) and chronosequence studies show that secondary forests mostly regenerate (\~50-80%, compared to old-growth forests) rather quickly (decades, \< 20 years) with respect to tree height, above-ground biomass (Poorter et al 2016), and also species richness in wetter regions (Rozendaal et al 2019, Onyekwelu & Olabiwonnu 2016). Higher richness may be enough to maintain carbon storage (Liu et al 2018), but lagging changes in composition may contribute to reduced effects of diversity on other forest functions (Griess et al 2012). Unlike richness, changes to species composition likely last up to centuries, showing a strong legacy of early-successional communities that are mostly fast-growing, low-wood-density, shade-intolerant pioneering species (Tabarelli et al 2014, Tabarelli & Lopes 2008). Slower changes in composition can be influenced by seed banks, herbivore communities, and dispersal from the surrounding landscape, which may also be managed, possibly explaining why different tropical forest trajectories diverge over time (Nordon et al 2015, Christensen 2014, Munroe et al 2013, Chazdon et al 2007).

Edge effects, or changes in composition along management gradients, are likely a key variable for the sustainability of neo-tropical landscapes that are heterogenous in management (Wright et al 2009), with underlying arguments based on metapopulation theory in ecology (Levins 1969, MacArthur & Wilson 1967), but edge effects have often been studied as threats to nature reserves, rather than how edges might facilitate secondary succession. Especially for abandoned land that regenerates via various processes (Stanturf et al 2014), the proximity to less disturbed habitat can affect the composition of the regrowing community by changing effective seed dispersal kernels and probability of establishment (Wandrag et al 2017, Guariguata & Ostertag 2001), along with animal disperser communities themselves (Gilroy et al 2014). While dispersal is complicated (Muller-Landau & Hardesty 2005), it can broadly correlate with simpler plant traits such as height and seed mass (Tamme et al 2014, Thomson et al 2011). Edge effects can also operate via the style of management, including logging decisions (Edwards et al 2014, Clark et al 2012) and land-use history generally (Omeja et al 2012), can also affect stand diversity maintenance, with examples of high diversity (compared to older adjacent stands) in agroforests (Oliveira-Neto et al 2017) and native timber plantations (Pryde et al 2015), possibly by changing early tree survival probabilities, growth rates, and/or competition, but surprisingly less likely facilitation by nitrogen-fixing trees (Lai et al 2018). It may also be that due to both high recruitment and mortality (turnover) of pioneering species' stems, transitions from early- to late-successional communities may be more complex than gradual (van Greugel et al 2007). The persistence of early-successional tree communities has been found to lower the standing biomass, likely due to lower wood density (César et al 2021), and could therefore require a switch in focal timber species (Karsten et al 2014).

Here we study the regeneration of an abandoned timber plantation as a function of distance from surrounding native forest, as means of understanding how these local ecological principles can support forest stewardship (Hulshoff & Powers 2019). We hypothesized the planted canopy species would facilitate regeneration by reducing competitive suppression of more shade-tolerant taxa by pioneering species, having mixed effects on stand diversity. We also considered how the legacy plantation canopy structure mediated forest stand effects via changes in abiotic conditions like light availability and microtopography (Powers & Peréz-Aviles 2012).

@@

```{=html}
<!-- -->
```
-   Neighborhood influence metric (Lorimer 1983, applied in Looney et al 2016)
-   Data and analytical code available on GitHub.../DRYAD

>>>>

Forested areas in the tropics offer many [economic and regulatory] services like timber, fuel, biomass/carbon storage, and biodiversity maintenance.

At the same time, most forested areas in the tropics are secondary (FAO 2020), showing legacy effects of human activity like logging and intensive agriculture.

As a result, tropical reforestation is a widespread ecological process that must be better understood to optimize [local] tradeoffs in ecosystem services [CITE?].

Previous studies have shown some general/consistent patterns of secondary forest succession, although other studies show inconsistent trajectories.

For example, syntheses (Finegan 1996) and chronosequence studies show that secondary forests mostly regenerate (\~50-80%, compared to old-growth forests) rather quickly (decades, \< 20 years) with respect to tree height, above-ground biomass (Poorter et al 2016), and species richness, especially in wetter regions (Rozendaal et al 2019, Onyekwelu & Olabiwonnu 2016).

However, while restoring species richness may be enough to maintain carbon storage (Liu et al 2018), it does not guarantee the restoration of species composition.

Instead, changes to species composition likely last up to centuries, showing a strong legacy of priority effects from early-successional communities that are mostly fast-growing, low-wood-density, shade-intolerant pioneering species (Tabarelli et al 2014, Tabarelli & Lopes 2008).

Slower changes in composition can also be influenced by seed banks, herbivore communities, and dispersal from the surrounding landscape, all of which may affect diverging restoration trajectories among tropical forests (Nordon et al 2015, Christensen 2014, Munroe et al 2013, Chazdon et al 2007).

but lagging changes in composition may contribute to reduced effects of diversity on other forest functions (Griess et al 2012).

Restoring species composition more quickly is likely a place where active forest managers can have the most unique impacts.

For land managers / foresters with interest in conservation, inheriting land affected by past decisions is a reality. One of the legacies is where the land was placed. To best manage it in the future, an initial assessment of the ecological status of the land is important to understand it's likely trajectory, and what it will need in the future for tailed services/conservation/protection.

Edge effects, or changes in composition along management gradients, are likely a key variable for the sustainability of neo-tropical landscapes that are heterogenous in management (Wright et al 2009),

with underlying arguments based on metapopulation theory in ecology (Levins 1969, MacArthur & Wilson 1967), but edge effects have often been studied as threats to nature reserves, rather than how edges might facilitate secondary succession.

Especially for abandoned land that regenerates via various processes (Stanturf et al 2014), the proximity to less disturbed habitat can affect the composition of the regrowing community by changing effective seed dispersal kernels and probability of establishment (Wandrag et al 2017, Guariguata & Ostertag 2001), along with animal disperser communities themselves (Gilroy et al 2014).

While dispersal is complicated (Muller-Landau & Hardesty 2005), it can broadly correlate with simpler plant traits such as height and seed mass (Tamme et al 2014, Thomson et al 2011).

Edge effects can also operate via the style of management, including logging decisions (Edwards et al 2014, Clark et al 2012) and land-use history generally (Omeja et al 2012), can also affect stand diversity maintenance,

with examples of high diversity (compared to older adjacent stands) in agroforests (Oliveira-Neto et al 2017) and native timber plantations (Pryde et al 2015),

possibly by changing early tree survival probabilities, growth rates, and/or competition,

but surprisingly less likely facilitation by nitrogen-fixing trees (Lai et al 2018).

It may also be that due to both high recruitment and mortality (turnover) of pioneering species' stems, transitions from early- to late-successional communities may be more complex than gradual (van Greugel et al 2007).

The persistence of early-successional tree communities has been found to lower the standing biomass, likely due to lower wood density (César et al 2021), and could therefore require a switch in focal timber species (Karsten et al 2014).

To understand the ecological principles behind secondary forest regeneration (Hulshoff & Powers 2019), here we study the regeneration of an abandoned timber plantation as a function of distance from surrounding native forest.

We hypothesized that planted canopy species would facilitate regeneration by reducing competitive suppression of more shade-tolerant taxa by pioneering species, having mixed effects on stand diversity.

We also considered how the legacy plantation canopy structure mediated forest stand effects via changes in abiotic conditions like light availability and microtopography (Powers & Peréz-Aviles 2012).

@@

-   Neighborhood influence metric (Lorimer 1983, applied in Looney et al 2016)
-   Data and analytical code available on GitHub.../DRYAD
